---
title: "Signal Four Analytics in Hillsborough County"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

Prep the workspace

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

```{r}
rm(list=ls())
```

Import libraries

```{r}
library(ggplot2)
library(lubridate)
library(dplyr)
library(caret)
```

load the data

```{r}
data = read.csv('data/crash_event.csv', sep=',', header=T)
head(data)
```

Add FDOT's peak period based on the crash date and time.

```{r}
data = data %>%
  mutate(CRASH_DATE_AND_TIME = dmy_hm(CRASH_DATE_AND_TIME)) %>%
  mutate(PEAK = as.factor(ifelse(
              between(as.numeric(format(CRASH_DATE_AND_TIME, "%H")), 7, 9) & !(weekdays(CRASH_DATE_AND_TIME) %in% c("Saturday", "Sunday")), "AM PEAK",
                  ifelse(
              between(as.numeric(format(CRASH_DATE_AND_TIME, "%H")), 10, 15) & !(weekdays(CRASH_DATE_AND_TIME) %in% c("Saturday", "Sunday")), "MID PEAK",
                    ifelse(
              between(as.numeric(format(CRASH_DATE_AND_TIME, "%H")), 16, 18) & !(weekdays(CRASH_DATE_AND_TIME) %in% c("Saturday", "Sunday")), "PM PEAK",
                     ifelse(
              between(as.numeric(format(CRASH_DATE_AND_TIME, "%H")), 6, 19) & (weekdays(CRASH_DATE_AND_TIME) %in% c("Saturday", "Sunday")), "WEEKEND PEAK",
            "OFF PEAK"
        )))))) %>%

  mutate(INCAPACITATING_FLAG = ifelse(
    S4_CRASH_SEVERITY %in% c('Fatality', 'Serious Injury'), 1, 0))

summary(data$INCAPACITATING_FLAG)
```

Filter to only relevant columns for analysis and ensure proper datatypes

```{r}
selected_columns <- c('INCAPACITATING_FLAG',
                      'PEAK',
                      'TOTAL_NUMBER_OF_VEHICLES',
                      'TOTAL_NUMBER_OF_PERSONS',
                      'RURAL_OR_URBAN',
                      'ROAD_SYSTEM_IDENTIFER',
                      'ROAD_SURFACE_CONDITION',
                      'S4_CRASH_TYPE_SIMPLIFIED',
                      'S4_DAY_OR_NIGHT',
                      'S4_IS_AGGRESSIVE_DRIVING',
                      'S4_IS_ALCOHOL_RELATED',
                      'S4_IS_CMV_INVOLVED',
                      'S4_IS_DISTRACTED',
                      'S4_IS_DRUG_RELATED',
                      'S4_IS_HIT_AND_RUN',
                      'S4_IS_INTERSECTION_RELATED',
                      'S4_IS_LANE_DEPARTURE_RELATED',
                      'S4_IS_SPEEDING_RELATED',
                      'S4_TRAILER_COUNT',
                      'S4_MOTORCYCLE_COUNT',
                      'S4_MOPED_COUNT',
                      'S4_BICYCLIST_COUNT',
                      'S4_AGING_DRIVER_COUNT',
                      'S4_TEENAGER_DRIVER_COUNT',
                      'S4_UNRESTRAINED_COUNT'
                      )
dataCrash = data %>%
  select(selected_columns) %>%
  mutate(across(c('RURAL_OR_URBAN',
                  'ROAD_SYSTEM_IDENTIFER',
                  'ROAD_SURFACE_CONDITION',
                  'S4_CRASH_TYPE_SIMPLIFIED',
                  'S4_DAY_OR_NIGHT',
                  'S4_IS_AGGRESSIVE_DRIVING',
                  'S4_IS_ALCOHOL_RELATED',
                  'S4_IS_CMV_INVOLVED',
                  'S4_IS_DISTRACTED',
                  'S4_IS_DRUG_RELATED',
                  'S4_IS_HIT_AND_RUN',
                  'S4_IS_INTERSECTION_RELATED',
                  'S4_IS_LANE_DEPARTURE_RELATED',
                  'S4_IS_SPEEDING_RELATED'), as.factor))
  
  

head(dataCrash)
```

```{r}
par(mfrow = c(2, 2))
# Crash type plots
ggplot(dataCrash, aes(x = S4_CRASH_TYPE_SIMPLIFIED)) +
  geom_bar(aes(fill = factor(INCAPACITATING_FLAG, levels = c(0, 1), labels = c("No", "Yes")))) +
  scale_fill_manual(values = c("No" = "blue", "Yes" = "red")) +
  labs(fill = "Incapacitating Event", x = "Crash Type", y = NULL) +
  ggtitle("Crash Severity by Type") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(subset(dataCrash, INCAPACITATING_FLAG == 1), aes(x = S4_CRASH_TYPE_SIMPLIFIED)) +
  geom_bar(fill = "red") +
  labs(x = "Crash Type", y = NULL) +
  ggtitle("Incapacitating Events by Type") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Peak plots
ggplot(dataCrash, aes(x = PEAK)) +
  geom_bar(aes(fill = factor(INCAPACITATING_FLAG, levels = c(0, 1), labels = c("No", "Yes")))) +
  scale_fill_manual(values = c("No" = "blue", "Yes" = "red")) +
  labs(fill = "Incapacitating Event", x = "Period", y = NULL) +
  ggtitle("Crash Severity by Peak Period") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(subset(dataCrash, INCAPACITATING_FLAG == 1), aes(x = PEAK)) +
  geom_bar(fill = "red") +
  labs(x = "Period", y = NULL) +
  ggtitle("Incapacitating Events by Peak Period") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

Split the data into train and test data by an 90/10 split and re-sample data to balance the response

```{r}
library(ROSE)
set.seed(123)
train_ind <- sample(seq_len(nrow(dataCrash)), size = floor(nrow(dataCrash)*0.9))

crashtrain = dataCrash[train_ind,]
crashtest = dataCrash[-train_ind,]

crashtrain_balanced = ROSE(INCAPACITATING_FLAG~., crashtrain, N=2000)$data
summary(crashtrain_balanced$INCAPACITATING_FLAG)
```

Logistic regression - full

```{r}
modellogit = glm(INCAPACITATING_FLAG~., crashtrain_balanced, family='binomial'(link='logit'))
summary(modellogit)
```

Logistic regression - stepwise with k=2 penalty degrees of freedom (AIC)

```{r}
logstep = step(glm(INCAPACITATING_FLAG~., data = crashtrain_balanced), trace = F)
summary(logstep)

predlogstep = ifelse(predict(logstep, crashtest[,-1], type="response" ) < 0.5, 0, 1)

clog = confusionMatrix(as.factor(predlogstep), as.factor(crashtest$INCAPACITATING_FLAG))
clog

paste0('Logit stepwise testing error: ', 1-clog$overall['Accuracy'])
```

Random Forest

```{r}
library(randomForest)
modelrf = randomForest(as.factor(INCAPACITATING_FLAG)~., crashtrain_balanced, proximity=T, importance=T)
predrf <- predict(modelrf, crashtest)
crf = confusionMatrix(predrf, as.factor(crashtest$INCAPACITATING_FLAG))
crf

paste0('Random Forest Test Error: ', 1-crf$overall['Accuracy'])
```

Boosting

```{r}
library(gbm)

modelgbm = gbm(INCAPACITATING_FLAG~., data=crashtrain_balanced, distribution = 'bernoulli', n.trees = 5000, shrinkage = 0.01, interaction.depth = 1, cv.folds = 10)
summary(modelgbm)

## Find the estimated optimal number of iterations shown in a dashed blue line and store as a variable
perf_gbm = gbm.perf(modelgbm, method="cv")

## prediction
gbmpred = ifelse(predict(modelgbm,newdata = crashtest, n.trees=perf_gbm1, type="response") < 0.5, 0, 1) 
cgbm = confusionMatrix(as.factor(gbmpred), as.factor(crashtest$INCAPACITATING_FLAG))
cgbm

paste0('GBM Test Error: ', 1-cgbm$overall['Accuracy'])
```
